### 2021-11-23 article.tex
Remove CodeSection suffixes
(Ab)Using Implicits

\SystemFm supports both pattern matching on the term level (\emph{match expressions}) as well as on the type-level (\emph{match types}).
Matches, both on terms and on types, are composed of a scrutinee, a list of cases and a default expression/type.
Each case consists of a \emph{type test} and a corresponding expression/type.
At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \citep{abadi1991dynamic}).
At the type level, a type test corresponds to a subtyping test with a particular type.
This disparity reflects the difference between runtime, where type tests are implemented using class tables, and compile time, where types are compared using the type system in its full extent.
We discuss the representation of Scala types at runtime in \Cref{subsec:types-at-runtime}.

@@

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \citep{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.

@@

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \citep{fengyun2016generic}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

@@

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \citep{schinz2005compiling}.
For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.

@@

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
%
\shapeCodeSection
%
This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.

To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
%
\ndarrayCodeSection

@@

To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalCodeSection

@@

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplyCodeSection
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

@@

An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementsCodeSection

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
%
\reshapeCodeSection
%
With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

@@

\npmeanCodeSection

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reduceCodeSection

@@

Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
%
\badreducescalaCodeSection
%
As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.

% \goodreducescalaCodeSection

>>>>

\SystemFm supports both pattern matching on the term level (\emph{match expressions}) as well as on the type-level (\emph{match types}).
Matches, both on terms and on types, are composed of a scrutinee, a list of cases and a default expression/type.
Each case consists of a \emph{type test} and a corresponding expression/type.
At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase~\citep{abadi1991dynamic}).
At the type level, a type test corresponds to a subtyping test with a particular type.
This disparity reflects the difference between runtime, where type tests are implemented using class tables, and compile time, where types are compared using the type system in its full extent.
We discuss the representation of Scala types at runtime in \Cref{subsec:types-at-runtime}.

@@

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null}~\citep{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.

@@

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses~\citep{fengyun2016generic}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

@@

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased~\citep{schinz2005compiling}.
For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.

@@

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
%
\shapeEnum
%
This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.

To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
%
\ndarrayTrait

@@

To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalDef

@@

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplyDef
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

@@

An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementsType

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
%
\reshapeDef
%
With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

@@

\npmeanDef

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reduceType

@@

Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
%
\badreducescala
%
As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.

% \goodreducescala
